---
title: "Group 7 — ERCP trial and the Americas"
subtitle: "Studio wall"
author: "Gandalf the Grey (Arvind) and 20 Peasants"
format:
  html:
    toc: true
    toc-depth: 2
code-tools: true
---

# Spine

> Does a simple pill change post-procedure pancreatitis, who is in that trial, and where in the Americas is life long?

~~This sentence stays at the top. Every panel must attach to it.~~

# Introduction — the tables we own

This folder is ours for the ten days. Three tables, three shapes.

- **Panel A** (`panel-A.csv`) — `medicaldata::indo_rct`: age and risk score, plus gender, treatment (`rx`: indomethacin vs placebo), and outcome.
- **Panel B** (`panel-B.csv`) — same trial, Quals: treatment, outcome, PEP, outpatient status, site.
- **Panel C** (`panel-C.csv`) — gapminder 2007, **Americas only**. Join by country name.

This is an RCT. `rx` is assigned, not a lifestyle. Do not talk about it as if people chose the pill.

```{r}
#| label: setup
#| message: false # keeps rendered Quarto less noisy
#| 
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
##
library(vcd) # (NEW!!) for Panel B
library(broom) # (NEW!!) for clean stat outputs


A <- read_csv("../../../data/group-7-ercp-and-americas/panel-A.csv")
B <- read_csv("../../../data/group-7-ercp-and-americas/panel-B.csv")
C <- read_csv("../../../data/group-7-ercp-and-americas/panel-C.csv")
```

~~Found tables train the mark. A campus Free Hunch is a **made** panel of the same pair type.~~

# Panel A — body / measure

> *age and risk are Quant. gender, rx, outcome are Quals for split.*


### 1. See the table

~~What is a case? Which columns are Quant / Qual? What is missing?~~

```{r}
#| label: inspect-panel-A-data-1
#| 
# glimpse() / inspect() / skim()
dplyr::glimpse(A)
```
My Panel-A data has:

  - 602 observations and 5 variables.
  - `age`: Quantitative, numeric
  - `risk`: Quantitative, numeric
  - `gender`: Qualitative, categorical. Encoded <chr>, to be mutated
  - `rx`: Qualitative, categorical. Encoded <chr>, to be mutated
  - `outcome`: Qualitative, categorical. Encoded <chr>, to be mutated

```{r}
#| label: inspect-panel-A-data-2
#| 
# glimpse() / inspect() / skim()
skimr::skim(A)
```


Furthermore, on the Qual variables panel-A data has : 

- `gender` has 2 levels: "male" and "female"
- `rx` has 2 levels: "indomethacin" and "placebo"
- `outcome` has 2 levels: "no" and "yes"

And on the Quant variables panel-A data has :

- `age` has a mean of 45.26, sd of 13.2, median of 45, min of 19, and max of 90
- `risk` has a mean of 2.38, sd of 0.88, median of 2.5, min of 1, and max of 5.5

There is no missing data. `age` histogram shown in `skim()` output seems skewed to the right; `risk` is slightly skewed right / normal. 

```{r}
#| label: mutate-A-factors

A_mod <- A |> 
  mutate(
    gender = as.factor(gender),
    rx = as.factor(rx),
    outcome = as.factor(outcome)
  )

glimpse(A_mod)

```
All ready now. 


### 2. Ask

~~One question *this* table can answer. One it cannot. Write both.~~
Looking at the overall Spine statement, I can pose the following Questions:

Possibly can answer: 

1. Are the dataset balanced by `gender` and by `rx` (assigned medication)?
1. What are the distributions for each of the Quant variables, `age`, and `risk`?
1. How are they different based on the Qual variables? (e.g., is there a difference in age distribution between `gender`-s, or between `rx` groups?)
1. Is the `risk` different for men and women? (`Main Character` Question)

Possibly cannot answer:

1. Does the medication efficacy depend upon dosage? ( No `dosage` variable.)
1. Better than some other medicine? ( No data, need an entire new table)


### 3. Choose a mark

~~Geom, map mark, or (later) test that matches the pair type. Name it before you code.~~

1. A bar chart
2. Histograms
3. Histograms, coloured by `gender` and `rx`
4. Histogram?? AND Stat Test For Means?? (NEW!!)

### 4. Predict

~~What do we think we will see? One sentence, before Run.~~

I think we will see that there is a difference in `risk` between men and women. 

And a Stat test for means will confirm this.

### 5. Make

~~Human brick first. Optional: one AI prompt, pasted below, only for what section 3 named.~~

These first few charts are what I want to check out. I may/will not have these charts in the final rendered Quarto, but want to comment on findings if needed. 

If any chart is surprising, it stays. 

```{r}
#| label: panelA-charts-I-want-to-see-but-not-show-1
#| eval: false
#| echo: true # Include the code, but don't compute in render

## Question 1
A_mod |> gf_bar(~ gender)

## Question 2
A_mod |> gf_bar(~ rx)

```

```{r}
#| label: panelA-charts-I-want-to-see-but-not-show-2
#| eval: false
#| echo: true # Include the code, but don't compute in render

## Question 3
A_mod |> gf_histogram(~ age)
A_mod |> gf_histogram(~ age, fill = ~ gender)
A |> gf_histogram(~ age, fill = ~ rx)
##
A_mod |> gf_histogram(~ risk)

```

```{r}
#| label: panel-A-Main-Character-question
## Question #4 (Main Character)

A_mod |> gf_histogram(~ risk, fill = ~ gender)
A_mod |> gf_density(~ risk, fill = ~ gender)

```

There are more `women` than `men` in the sample. 

Distributions do change with `gender`. But hard to say outright from the overlapping histograms. *Densities* for `risk` look different for men and women, though. 


```{r}
#| label: panel-A-stat-test
#| echo: true

# Main Character Question

mosaic::t.test(risk ~ gender, data = A_mod) |> broom::tidy()

```

### 6. Say

~~Caption + one sentence we will stand behind.~~


1. Women are literally at higher `risk` than men for post-procedure pancreatitis, no cap. It’s giving sad energy. 
2. Difference in mean `risk` between them is $0.3014706$, statistically significant (p-value < 0.05).

### 7. Limit

~~What this panel cannot do.~~

We’re assuming the community here is not specifically linked with pancreatitis. Would this `risk` be similar in other communities? (Lowkey assuming this community isn’t pancreatitis-coded. Would the risk vibe be similar in other communities, no cap?)

# Panel B — category / talk

*rx × outcome is the trial mosaic. site × outcome. Do not average rx.*


### 1. See the table

~~What is a case? Which columns are Quant / Qual? What is missing?~~

```{r}
#| label: inspect-panel-B-data-1
#| 
# glimpse() / inspect() / skim()
dplyr::glimpse(B)
```
My Panel-B data has:

- 602 observations and 5 variables.
- `rx`: Qualitative, categorical. Encoded <chr>, to be mutated
- `outcome`: Qualitative, categorical. Encoded <chr>, to be mutated
- `pep`: Qualitative, categorical. Encoded <chr>, to be mutated
- `status`: Qualitative, categorical. Encoded <chr>, to be mutated
- `type`: Qualitative, categorical. Encoded <chr>, to be mutated
- `site`: Qualitative, categorical. Encoded <chr>, to be mutated

```{r}
#| label: inspect-panel-B-data-2
#| 
# glimpse() / inspect() / skim()
skimr::skim(B)
```

Further, on the Qual variables panel-B data has : 

- `gender` has 2 levels: "male" and "female"
- `rx` has 2 levels: "indomethacin" and "placebo"
- `outcome` has 2 levels: "no" and "yes"
- `pep` has 2 levels: "no" and "yes"
- `status` has 2 levels: "outpatient" and "inpatient"
- `type` has 4 levels: to be listed
- `site` has 4 levels: to be listed

And on the Quant variables panel-A data has : None. All variable are Qualitative only. 

There is no missing data.

```{r}
#| label: mutate-B-factors

B_mod <- B |> 
  mutate(across(where(is.character), as.factor)) # Whoa!!

glimpse(B_mod)

```

Now, we're ready.

### 2. Ask

~~One question *this* table can answer. One it cannot. Write both.~~

Looking at the spine statement again, I can pose the following Questions:

Possibly can answer:

1. Are the dataset balanced by all Qual variables, except the measured one (`output`) ?
1. Are there differences in *how the medication was administered* ( i.e.`rx`)  across `gender`, `pep`, `site` or `type` of patient? ( Just curious. Does this matter? Need to ask the expert who did this research.)
1. Does the medication work? (Main Character Question)
1. Does it work differently for men and for women? ( Possible Main Character Question)

Possibly cannot answer:

1. Does the medication efficacy depend upon dosage? ( No `dosage` variable.)

### 3. Choose a mark

1. Bar Chart
1. Bar Chart with fill or facet
1. Mosaic Chart (NEW!!)
1.  Mosaic Chart again??

### 4. Predict

~~What do we think we will see? One sentence, before Run.~~

A well-balanced RCT experiment, balanced across all `factors` such as `gender`, `pep`, `site`, and `type`. 

I think the medicine might show as working in this experiment. 

### 5. Make

~~Human brick first. Optional: one AI prompt, pasted below, only for what section 3 named.~~

These first few charts are what I want to check out. I will not have these charts in the final rendered Quarto, but comment on findings. 

If any is surprising, it stays. 

```{r}
#| label: panel-B-charts-I-want-to-see-but-not-show-1
#| eval: false
#| echo: true # Include the code, but don't compute in render

## Question 1
B_mod |> gf_bar(~ gender)
B_mod |> gf_bar(~ rx)
B_mod |> gf_bar(~ pep)
B_mod |> gf_bar(~ status)
B_mod |> gf_bar(~ type)
B_mod |> gf_bar(~ site)

```
- `gender`: More women than men
- ` rx`: Medication and placebo prescribed to equal numbers. Fair. 
- All other factors have imbalanced counts. It is what it is. 
```{r}
#| label: panel-B-charts-I-want-to-see-but-not-show-2
#| eval: false
#| echo: true # Include the code, but don't compute in render

## Question 2
B_mod |> gf_bar(~ gender, fill = ~ rx)
B_mod |> gf_bar(~ pep, fill = ~ rx)
B_mod |> gf_bar(~ status, fill = ~ rx)
B_mod |> gf_bar(~ type, fill = ~ rx)
B_mod |> gf_bar(~ site, fill = ~ rx)

```
All kinds of people, based on ***each of the factors*** are EQUALLY divided between `indomethacin` and `placebo`. Superb experiment design!! Waah, waah!!

NOTE: this can be visually confirmed by using `gf_props` for all code lines in the preceding chunk. 


```{r}
#| label: main-character-question-B
#| echo: true

# Question #3
# library(vcd) # Put this in the setup chunk above!
# 
B_table <- vcd::structable(outcome ~ rx, data = B_mod)
B_table
# Main Character Question
vcd::mosaic(B_table, shade = TRUE, main = "Mosaic Chart for Indomethacin RCT")
```


### 6. Say

~~Caption + one sentence we will stand behind.~~

The mosaic plot’s saying there’s a significant diff between `indomethacin` and `placebo` performance. But plot twist: the diff is wrong-sided: placebo’s actually slaying. Main character `indomethacin` is kinda flopping compared to placebo. Low-key mid, high-key not it. We need a new main. 

### 7. Limit

~~What this panel cannot do.~~

I cannot say whether this trial would show similar results if conducted in another community. I presume the experimenters thought of this beforehand. 


# Panel C — place

*Americas-only lifeExp / gdpPercap / pop. Same join brick as other map panels.*

```r
# library(sf)
# library(rnaturalearth)
# world <- ne_countries(scale = "medium", returnclass = "sf")
```


### 1. See the table

What is a case? Which columns are Quant / Qual? What is missing?

```r
# glimpse() / inspect() / skim()
```

### 2. Ask

One question *this* table can answer. One it cannot. Write both.

### 3. Choose a mark

Geom, map mark, or (later) test that matches the pair type. Name it before you code.

### 4. Predict

What do we think we will see? One sentence, before Run.

### 5. Make

Human brick first. Optional: one AI prompt, pasted below, only for what section 3 named.

```r
# human chunk
```

### 6. Say

Caption + one sentence we will stand behind.

### 7. Limit

What this panel cannot do.



# Conclusion

- The question we will defend in ninety seconds
- What we will **not** claim
- Which lines were human, which were machine
- Limits
- Gift for the steal-this strip
