---
title: "Group 2 — Risk factors"
subtitle: "Studio wall"
author: "Group names here"
format:
  html:
    toc: true
    toc-depth: 2
---

# Spine

> How do bodies carry risk, how do class and smoke talk, and where do children still die young?

This sentence stays at the top. Every panel must attach to it. If a plot cannot be introduced in a clause that hangs off this line, it is off the wall.

# Introduction — the tables we own

This folder is ours for the ten days. Three tables, three shapes.

- **Panel A** (`panel-A.csv`) — `cardioRiskFactors`: BMI, lipids, blood pressure, uric acid. `sex` and `smok` have already been recoded from integers to factors.
- **Panel B** (`panel-B.csv`) — Scottish Health Survey 1998: sex, social class, smoking, diagnosed CVD. `area` is a code, not a map.
- **Panel C** (`panel-C.csv`) — SOWC child / infant / neonatal mortality by country. Join by country name.

If a brick needs a histogram, it lives in A. If it needs a mosaic, it lives in B.

We load from this folder:

```{r}
#| label: setup
#| message: false # keeps rendered Quarto less noisy
#| 
library(tidyverse)
library(mosaic)
library(skimr)
library(ggformula)
##
library(vcd) # for Panel B


A <- read_csv("../../../data/group-2-risk-factors/panel-A.csv")
B <- read_csv("../../../data/group-2-risk-factors/panel-B.csv")
C <- read_csv("../../../data/group-2-risk-factors/panel-C.csv")
```

~Found tables train the mark. A campus Free Hunch, if we collect one, is a **made** panel of the same pair type — new people, old grammar. It does not replace A, B, or C.~

# Panel A — body / measure

*bmi, choles, sys, dia, Uric are Quant. Do not average `smok`.*


### 1. See the table

What is a case? Which columns are Quant / Qual? What is missing?

```{r}
#| label: inspect-panel-A-data-1
#| 
# glimpse() / inspect() / skim()
dplyr::glimpse(A)
```

```{r}
A_mod <- A |> mutate(
  sex = as.factor(sex),
  smok = as.factor(smok)) |> 
  
  select(sex, smok, everything()) # factors first

glimpse(A_mod)
```
My Panel-A data has:

  - 998 observations and 11 variables
  - `smok`, `sex` are Qual variables
  - `bmi`, `waisthip`, `choles`, `ldl`, `hdl`, `sys`, `dia`, `Uric`, `alco`, and `apoa` are Quant variables
  
### 2. Ask

~~One question *this* table can answer. One it cannot. Write both.~~

Looking at the spine statement, I can pose the following Questions:

Possibly can answer: 

1. Are the dataset balanced by `sex` and by `smok` ?
1. What are the distributions for each of the Quant variables?
1. How are they different based on the Qual variables? Is there any risk ( Quant variable) that is different based on `sex` or `smok`?

Possibly cannot answer:

1. There is no data on disease or other symptoms, so cannot plot that. 

### 3. Choose a mark

~Geom, map mark, or (later) test that matches the pair type. Name it before you code.~

- Bar Charts
- Histograms
- Histograms with fill
- Boxplots for variety

### 4. Predict

What do we think we will see? One sentence, before Run.

### 5. Make

Human brick first. Optional: one AI prompt, pasted below, only for what section 3 named.

```{r}
#| label: panelA-charts-I-want-to-see-but-not-show-1
#| eval: false
#| echo: true # Include the code, but don't compute in render

## Question 1
A_mod |> gf_histogram(~ bmi)
A_mod |> gf_histogram(~ waisthip)
A_mod |> gf_histogram(~ choles)
A_mod |> gf_histogram(~ ldl)
A_mod |> gf_histogram(~ hdl)
A_mod |> gf_histogram(~ sys)
A_mod |> gf_histogram(~ dia)
A_mod |> gf_histogram(~ Uric)
A_mod |> gf_histogram(~ alco)
A_mod |> gf_histogram(~ apoa)

## Question 2
A_mod |> gf_histogram(~ choles, fill = ~ sex)
A_mod |> gf_histogram(~ choles, fill = ~ smok)
A_mod |> gf_histogram(~ ldl, fill = ~ sex)
A_mod |> gf_histogram(~ ldl, fill = ~ smok)
A_mod |> gf_histogram(~ hdl, fill = ~ sex)
A_mod |> gf_histogram(~ hdl, fill = ~ smok)
A_mod |> gf_histogram(~ sys, fill = ~ sex)
A_mod |> gf_histogram(~ sys, fill = ~ smok)
A_mod |> gf_histogram(~ dia, fill = ~ sex)
A_mod |> gf_histogram(~ dia, fill = ~ smok)
A_mod |> gf_histogram(~ alco, fill = ~ sex)
A_mod |> gf_histogram(~ alco, fill = ~ smok)
A_mod |> gf_histogram(~ apoa, fill = ~ sex)
A_mod |> gf_histogram(~ apoa, fill = ~ smok)
## Just for variety
A_mod |> gf_boxplot(sex ~ apoa, fill = ~ sex)
A_mod |> gf_boxplot(smok ~ apoa, fill = ~ smok)

```

```{r}
A_mod |> gf_histogram(~ Uric, fill = ~ sex)
A_mod |> gf_histogram(~ Uric, fill = ~ smok)
```

### 6. Say

`Uric` acid distributions look wild! There is so much right-skew to that distribution; the max value is near 2000! Someone suffers from a serious case of gout!


### 7. Limit

~What this panel cannot do. Which kit table we refused to abuse.~

Can only speak of general conditions and their risks. Cannot actually say that the risk has manifested!


# Panel B — category / talk

*smoke × sex, cvd × sc (social class). `area` is forbidden as geography.*


### 1. See the table

What is a case? Which columns are Quant / Qual? What is missing?

```{r}
#| label: inspect-panel-B-data-1
#| 
# glimpse() / inspect() / skim()
dplyr::glimpse(B)
```

```{r}
B_mod <- B |> 
  mutate(across(where(is.character), as_factor)) |> 
  select(sex, cvd, smoke, everything()) # factors first

glimpse(B_mod)
B |> count(smoke) # What are the levels?
```

My Panel-B data has:

- 8804 observations with 7 variables
- `sex`, `cvd`, `smoke` are Qual variables
-  `age` is Quant variable ( Age in years)
- `sc` is social class, encoded as Quant
- `carstair` is Quant variable, meaning not clear, perhaps related to exercise

```{r}
#| label: inspect-panel-B-data

skimr::skim(B_mod)

```
Further, on the Qual variables panel-B data has : 

- `sex` has 2 levels: "mal" and "fem"
- `cvd` has 2 levels: "no" and "yes"
- `smoke` has 3 levels: "nev", "lite", "mod", "hvy", "ex". Need to make this into an ordered factor!

And on the Quant variables panel-B data has :

- `age` has a mean of 45.26, sd of 13.2, median of 45, min of 19, and max of 90
- `sc` has a mean of 18.34, sd of 18.092, median of 12, min of 3, and max of 45
- `carstair` has a mean of -0.20, sd of 3.26, median of -0.68, min of -6.23, and max of 12.53
- `area` has a mean of 162, sd of 91, median of 165, min of 1, and max of 312. 

There is no missing data. All histograms look uniform!!

```{r}
B_mod <- B_mod |> 
  mutate(smoke = base::factor(smoke, 
                              levels = c("nevr", "lite", "mod", "hvy", "ex"), 
                              ordered = TRUE))
glimpse(B_mod)
```


### 2. Ask

~~One question *this* table can answer. One it cannot. Write both.~~
Looking at the spine statement, I can pose the following Questions:

Possibly can answer: 

1. Are the dataset balanced by `sex`, `cvd`, and `smoke`? 
1. Are `cvd` and `smoke` associated? (Main Character Question)
1. Are `cvd` and `sc` associated ? ( Also Main Character Question)

Possibly cannot answer:

1. Whether `smoke`-ing CAUSES `cvd`? Why not??

### 3. Choose a mark

~~Geom, map mark, or (later) test that matches the pair type. Name it before you code.~~

1. Bar charts
2. Mosaic Charts(NEW!)

### 4. Predict

~What do we think we will see? One sentence, before Run.~
I expect to see `cvd` worsen with `smoke`ing. `sc` may be curiously associated with `cvd` but may be confounded by `smoke`ing.


### 5. Make

~~Human brick first. Optional: one AI prompt, pasted below, only for what section 3 named.~~

```{r}
#| label: panelB-charts-I-want-to-see-but-not-show-1
#| eval: false
#| echo: true # Include the code, but don't compute in render

## Question 1
B_mod |> gf_bar(~ cvd)
B_mod |> gf_bar(~ sc)
B_mod |> gf_bar(~ sex)
```

`sc` is not really balanced at all, nor is `sex`.

`cvd` is not balanced, but it is an output variable. 

```{r}
#| label: main-character-question-B
#| echo: true

# Question #2
sccvd_table <- vcd::structable(sc ~ cvd, data = B_mod)
sccvd_table
smokecvd_tab <- vcd::structable(smoke ~ cvd, data = B_mod)
smokecvd_tab

##
vcd::mosaic(sccvd_table, shade = TRUE, legend = TRUE, main = "Mosaic of social class vs CVD", sub = "Actual")
vcd::mosaic(sccvd_table, shade = TRUE, legend = TRUE, main = "Mosaic of social class vs CVD", sub = "Expected", type = "expected")
vcd::mosaic(smokecvd_tab, shade = TRUE, legend = TRUE, main = "Mosaic of smoking vs CVD", sub = "Actual")
vcd::mosaic(smokecvd_tab, shade = TRUE, legend = TRUE, main = "Mosaic of smoking vs CVD", sub = "Expected", type = "expected")
```

### 6. Say

~~Caption + one sentence we will stand behind. Title is a sentence, not a column name.~~

Smoking and Cardio-Vascular Disease are definitely associated, with a p-value of <2.22e-16!!

### 7. Limit

~~What this panel cannot do. Which kit table we refused to abuse.~~

Can't PROVE that smoking causes cancer, since this is not an RCT Table!!

# Panel C — place

*under5_mortality_2018 and infant_mortality_2018 are rates. Join on `country`.*

Geometry is not in the CSV. Join in R, then use the same seven headings.

```r
# library(sf)
# library(rnaturalearth)
# world <- ne_countries(scale = "medium", returnclass = "sf")
# # join on country name; list the unmatched labels
```


### 1. See the table

What is a case? Which columns are Quant / Qual? What is missing?

```r
# glimpse() / inspect() / skim()
```

### 2. Ask

One question *this* table can answer. One it cannot. Write both.

### 3. Choose a mark

Geom, map mark, or (later) test that matches the pair type. Name it before you code.

### 4. Predict

What do we think we will see? One sentence, before Run.

### 5. Make

Human brick first. Optional: one AI prompt, pasted below, only for what section 3 named.

```r
# human chunk

# ## Prompt we actually used
# ai_draft
# ours
```

### 6. Say

Caption + one sentence we will stand behind. Title is a sentence, not a column name.

### 7. Limit

What this panel cannot do. Which kit table we refused to abuse.


# Conclusion

What the spine looks like after three panels.

- The question we will defend in ninety seconds
- What we will **not** claim
- Which lines were human, which were machine
- Limits: sample, missingness, the test we did not run
- Gift for the steal-this strip (palette, annotation, join trick)

```r
# sessionInfo()
```
